Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-31 paper2.qmd
section:4.2; process:write

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear in the rest of the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$. Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.

@@

<!-- 3. how de we use that data structure? -->
Now, using equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors capture all the observations from the conceptual population. Specifically, $O^{cp}_{R}$ and $D_{R}$ are observed and latent vectors of length $n$, respectively. Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), and when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$. It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$. Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$. Similarly, $X_{IA}$ is an observed matrix with dimension $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$. It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.

@@

Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] define the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model. Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.

@@

Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$. In other words, researchers can only observe texts and judgments from students and judges initially included in the sample. Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$. Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all judges $(n^{s}_{J} \cdot n^{s}_{K})$.

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.

@@

The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32]. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons a subset of judges makes for a subset of stimuli produced by a subset of individuals* from the conceptual-population, ultimately generating the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

@@

However, because CJ experiments rarely exhaustively pair sampled judgments, judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p. 5]--a realistic scenario must account for judges comparing only specific stimuli from certain individuals.

>>>>

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear in the rest of the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.

@@

<!-- 3. how de we use that data structure? -->
Now, using equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors capture all the observations from the conceptual population. Specifically, $O^{cp}_{R}$ and $D_{R}$ are observed and latent vectors of length $n$, respectively. Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability. Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010]. Their hierarchical nature enables researchers to test hypotheses considering the hierarchical structure of stimuli (see @sec-theory-issue1b) and the uncertainties in trait estimation (see @sec-theory-issue2).

<!-- 2 Explaining what we did -->
The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$. It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$. Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$. Similarly, $X_{IA}$ is an observed matrix with dimension $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$. It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.

@@

Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model. Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.

@@

Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$. In other words, researchers can only observe texts and judgments from students and judges initially included in the sample. Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$. Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals from the population.

@@

The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32], in this case $S$. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the subset of individuals* sampled from the conceptual-population. This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

@@

However, because CJ experiments rarely use the exhaustive pairing of the sampled judgments, judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p. 5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
